---
title: "Interactive Plots"
author: "Matt Eldridge"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

In this section we briefly introduce creating Shiny applications with interactive plots. These are figures where clicking on a point within a plot will trigger a selection event that causes an update to another part of the application.

We will show how this can be achieved with static plots generated by ggplot2 and also with HighCharts which allows for even more interactivity including zooming and tooltips.

For the examples shown here we use the results of a differential expression analysis using [limma](https://bioconductor.org/packages/release/bioc/html/limma.html), comparing ER negative and positive status groups.

### ggplot2 figures

The following Shiny application shows a Volcano plot of the log p-value versus the log fold change.

Here we add a `click` argument to `plotOutput` to listen to mouse click events. You can also listen to double-click and hover (mouse-over) events in exactly the same way.

Note the way in which the output text box is updated in response to a mouse click reflected in a change to the value of `input$volcanoPlotClick`. `nearPoints` is a convenient function for extracting the rows of the data frame used to generate the plot for the points near where the mouse was clicked.

```{r message = FALSE}
library(shiny)
library(tidyverse)

differentialExpressionResults <-
  read.csv("NKI-DE-results.csv") %>%
  mutate(minusLog10Pvalue = -log10(adj.P.Val))

ui <- fluidPage(
  fluidRow(
    column(
      width = 6,
      plotOutput("volcanoPlot", click = "volcanoPlotClick")
    ),
    column(
      width = 6,
      verbatimTextOutput("info")
    )
  )
)

server <- function(input, output) {

  output$volcanoPlot <- renderPlot({
    differentialExpressionResults %>%
      ggplot(aes(x = logFC, y = minusLog10Pvalue)) +
      geom_point() +
      xlab("log fold change") +
      ylab("-log10(P-value)")
  })

  output$info <- renderPrint({
    nearPoints(differentialExpressionResults, input$volcanoPlotClick) %>%
      transmute(
        Probe = probe,
        Gene = HUGO.gene.symbol,
        logFC = signif(logFC, digits = 2),
        adj.P.Val = signif(adj.P.Val, digits = 2)
      ) %>%
      remove_rownames()
  })
}

shinyApp(ui, server)
```

Note that if you are using base graphics the `nearPoints` function needs to be provided with details of what columns were used for the x- and y-axes; for ggplot2 Shiny was able to infer these and so we didn't need to specify the `xvar` and `yvar` arguments.

#### Exercise

Try changing the code above to handle mouse-over (hover) and brush events (brushing involves drawing a box in the plot area) instead of mouse clicks.


